---
title: "2025 Data Science Task"
author: "TERESIA MWAGONA"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
---

```{r,echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)    
library(ggplot2)      
library(sf)           
library(lme4)
library(plotly)
library(dplyr)
library(stringr)
library(viridis)
library(scales)
# Load  the dataset 
hiv_data <- read_csv("HIV data 2000-2023.csv",, show_col_types = FALSE)

```

**Cleaning the HIV 2000_2023 DATASET**

```{r,echo=TRUE, message=FALSE, warning=FALSE,error=FALSE}

# Cleaning the data by extracting main estimate from value column
hiv_data <- hiv_data %>%
  mutate(
    HIV_Population_raw = str_extract(Value, "^[^\\[]+"),              
    HIV_Population_clean = str_replace_all(HIV_Population_raw, "\\s", ""),  # remove spaces
    HIV_Population = as.numeric(HIV_Population_clean)                # convert to number
  )
#Check rows that failed to convert
problematic_rows <- hiv_data %>%
  filter(is.na(HIV_Population) & !is.na(Value))

#Remove no data rows and < estimates
hiv_data <- hiv_data %>%
  filter(!str_detect(Value, "No data|<")) %>%  
  mutate(
    HIV_Population_raw = str_extract(Value, "^[^\\[]+"),
    HIV_Population_clean = str_replace_all(HIV_Population_raw, "\\s", ""),
    HIV_Population = as.numeric(HIV_Population_clean)
  ) %>%
  filter(!is.na(HIV_Population))  # Keep only valid rows

#drop last 2 columns
hiv_data <- hiv_data %>%
  select(-HIV_Population_raw, -HIV_Population_clean)

#rename hiv_clean data as Value
hiv_data <- hiv_data %>%
  select(-Value) %>%  # remove old Value column
  rename(Value = HIV_Population)  # rename cleaned column to Value

```

**Trends of HIV cases in countries that contribute to 75% of the global burden**

```{r,echo=TRUE, message=FALSE, warning=FALSE}

hiv_data %>%
  group_by(Period) %>%
  summarise(total = sum(Value, na.rm = TRUE)) %>%
  arrange(desc(Period))

# Sum the total HIV cases for each country across all period
total_hiv_by_country <- hiv_data %>%
  filter(!is.na(Value)) %>%  
  group_by(Location) %>%
  summarise(total_hiv = sum(Value, na.rm = TRUE)) %>%
  arrange(desc(total_hiv))

# Calculate cumulative sum and cumulative percentage of global HIV burden
total_hiv_sum <- sum(total_hiv_by_country$total_hiv)  

total_hiv_by_country <- total_hiv_by_country %>%
  mutate(
    cum_sum = cumsum(total_hiv),  # Cumulative sum of HIV cases
    cum_percent = cum_sum / total_hiv_sum 
  )

# Identify countries that contribute to 75% of the global HIV burden
top75_countries <- total_hiv_by_country %>%
  filter(cum_percent <= 0.75)%>%
  pull(Location)

# Print the countries contributing to 75% of the HIV burden
top75_countries

```

```{r,echo=TRUE, message=FALSE, warning=FALSE,fig.width=14, fig.height=8}
# filter data for the countries contributing to 75% of the HIV burden
hiv_trend_data <- hiv_data %>%
  filter(Location %in% top75_countries, !is.na(Value))

#plot the trends
ggplot(hiv_trend_data, aes(x = Period , y = Value, color = Location)) +
  geom_line() +
  geom_point() +
  facet_wrap(~Location, scales = "free_y") +   
  labs(
    title = "HIV Cases Trend by Country (75% Global HIV Burden)",
    x = "Period",
    y = "Number of People Living with HIV"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

**Trends of HIV Cases in Countries contributing 75% of the burden within each WHO region**

```{r,echo=TRUE, message=FALSE, warning=FALSE,fig.width=14, fig.height=12}
# Filter out missing values
hiv_clean <- hiv_data %>%
  filter(!is.na(Value))

#Identify top 75% burden countries WITHIN EACH REGION
top75_by_region <- hiv_clean %>%
  group_by(ParentLocationCode, Location) %>%
  summarise(total_hiv = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  group_by(ParentLocationCode) %>%
  arrange(desc(total_hiv), .by_group = TRUE) %>%
  mutate(
    cum_sum = cumsum(total_hiv),
    region_total = sum(total_hiv),
    cum_percent = cum_sum / region_total
  ) %>%
  filter(cum_percent <= 0.75) %>%
  select(ParentLocationCode, Location)

#Join with original data to get time trend info for these countries
hiv_top75 <- hiv_clean %>%
  inner_join(top75_by_region, by = c("ParentLocationCode", "Location"))

#Plot trends for each region
ggplot(hiv_top75, aes(x = Period, y = Value, color = Location)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ParentLocationCode, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  labs(
    title = "HIV Trends (2000–2023) in Top 75% Burden Countries Within Each WHO Region",
    x = "Year",
    y = "Estimated People Living with HIV",
    color = "Country"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")+
  theme(legend.text = element_text(size = 14))

```

**Analyzing the relationship between people living with HIV and multidimensional poverty and the individual factors that contribute to the ratio**

```{r,echo=TRUE, message=FALSE, warning=FALSE}
#Load the dataset
poverty_data <- readxl::read_excel("multidimensional_poverty.xlsx")  

# Extract the first row as the initial column names for columns 1 to 10 
initial_column_names <- poverty_data[1, 1:10]
# Extract the second row for columns 11 to 15
second_row_column_names <- poverty_data[2, 11:15]
# Extract the name of column 16 from the first row
column_16_name <- poverty_data[1, 16]
# Combine all parts of the column names
new_column_names <- c(as.character(initial_column_names), as.character(second_row_column_names), as.character(column_16_name))
# Assign the new column names to the data
names(poverty_data) <- new_column_names
names(poverty_data)

# Skip the first two rows
poverty_data <- poverty_data[-c(1, 2), ]

# Rename poverty columns
poverty_data <- poverty_data %>%
  rename(
    Location = Economy,
    Period = `Reporting year`
  )
# convert period in both data sets to numeric
hiv_data$Period <- as.numeric(hiv_data$Period)
poverty_data$Period <- as.numeric(poverty_data$Period)
```

```{r,echo=TRUE, message=FALSE, warning=FALSE}
# Merge datasets
merged_data <- hiv_data %>%
  select(Location, Period, Value,ParentLocationCode) %>%
  inner_join(poverty_data, by = c("Location", "Period"))
#renaming columns
merged_data <- merged_data %>%
  rename(
    Income = `Deprivation rate (share of population)`,
    Year = Period,
    Country = Location
  )
names(merged_data)
# Replace "-" with NA in the entire dataset
merged_data[merged_data == "-"] <- NA

# Convert the  variables to numeric 
merged_data$`Multidimensional poverty headcount ratio (%)` <- as.numeric(merged_data$`Multidimensional poverty headcount ratio (%)`)
merged_data$Income <- as.numeric(as.character(merged_data$Income))  
merged_data$`Educational attainment (%)` <- as.numeric(as.character(merged_data$`Educational attainment (%)`))  
merged_data$`Educational enrollment (%)` <- as.numeric(as.character(merged_data$`Educational enrollment (%)`))  
merged_data$`Electricity (%)` <- as.numeric(as.character(merged_data$`Electricity (%)`))  
merged_data$`Sanitation (%)` <- as.numeric(as.character(merged_data$`Sanitation (%)`))  
merged_data$`Drinking water (%)` <- as.numeric(as.character(merged_data$`Drinking water (%)`))  

# Standardize the predictors  
merged_data$Value <- scale(merged_data$Value)  
merged_data$Income <- scale(merged_data$Income)  
merged_data$`Educational attainment (%)` <- scale(merged_data$`Educational attainment (%)`)  
merged_data$`Educational enrollment (%)` <- scale(merged_data$`Educational enrollment (%)`)  
merged_data$`Electricity (%)` <- scale(merged_data$`Electricity (%)`)  
merged_data$`Sanitation (%)` <- scale(merged_data$`Sanitation (%)`)  
merged_data$`Drinking water (%)` <- scale(merged_data$`Drinking water (%)`)  
```

```{r,echo=TRUE, message=FALSE, warning=FALSE}

# Fit a mixed model
 mixed_model <- lmer(`Multidimensional poverty headcount ratio (%)` ~   
                       Value +   
                       Income +   
                        `Educational attainment (%)` +   
                       `Educational enrollment (%)` +   
                       `Electricity (%)` +   
                        `Sanitation (%)` +   
                       `Drinking water (%)` +   
                       (1 | Year),   
                     data = merged_data)  
 
# View the model summary  
 summary(mixed_model)
```

The analysis shows that higher HIV prevalence is linked to lower multidimensional poverty (β = -1.91), meaning countries with more people living with HIV tend to have less poverty.However, low income levels strongly increase poverty (β = 12.83), confirming that poorer countries face worse living conditions. Both school enrollment (β = 3.27, p \< 0.001) and electricity access (β = 5.04, p \< 0.001) significantly reduce poverty,showing that education and infrastructure matter.In contrast, educational attainment (%), sanitation access, and drinking water access did not show statistically significant effects (all p \> 0.05).The Random effect Year (SD = 6.80) suggests that poverty changes over time.It indicates unexplained yearly variations in poverty.

**Filter the data for the 8 countries belonging in EAC**

```{r,echo=TRUE, message=FALSE, warning=FALSE,error=FALSE}
mortality_dataset <- read_csv("dataset_datascience.csv",show_col_types = FALSE)  

u5_mortality_rate <- mortality_dataset %>% filter(Indicator=="Under-five mortality rate")
neonatal_rate <- mortality_dataset %>%  filter(Indicator=="Neonatal mortality rate")

#Define EAC countries
eac_countries <- c("Burundi", "Democratic Republic of the Congo", "Kenya", 
                   "Rwanda", "Somalia", "South Sudan", "Uganda", "United Republic of Tanzania")

#Filter the data for the 8 countries
under_five_eac <- u5_mortality_rate %>% filter(`Geographic area` %in% eac_countries)
neonatal_eac <- neonatal_rate %>% filter(`Geographic area` %in% eac_countries)

```

**Map for latest estimate of Under 5 Mortality rates**

```{r,echo= TRUE, message=FALSE, warning=FALSE,fig.width=10, fig.height=6}
#extracting the latest year 
under_five_eac <- under_five_eac %>%
  mutate(
    Year = ifelse(
      str_detect(`Series Year`, "-"),  
      str_sub(`Series Year`, -4, -1),  
      `Series Year`                    
    ),
    Year = as.numeric(Year)           
  )

neonatal_eac <- neonatal_eac %>%
  mutate(
    Year = ifelse(
      str_detect(`Series Year`, "-"),  
      str_sub(`Series Year`, -4, -1),  
      `Series Year`                    
    ),
    Year = as.numeric(Year)           
  )
# Pick the latest year
mortality_latest <- under_five_eac %>%
  filter(`Geographic area` %in% c("Kenya", "Rwanda", "Uganda", "United Republic of Tanzania", "South Sudan", "Burundi", "Democratic Republic of the Congo", "Somalia")) %>%
  group_by(`Geographic area`) %>%
  slice_max(order_by = Year, n = 1) %>%
  ungroup()

# 1. Filter to Sex == "Total" and Wealth Quintile == "Total"
mortality_latest <- mortality_latest %>%
  filter(Sex == "Total", `Wealth Quintile` == "Total")

#Get the median for the observation values
mortality_latest <- mortality_latest %>%
  group_by(`Geographic area`, Year) %>%
  summarise(Estimate = median(`Observation Value`, na.rm = TRUE)) %>%
  ungroup()

# Pick latest year
neonatal_latest <- neonatal_eac %>%
  filter(`Geographic area` %in% c("Kenya", "Rwanda", "Uganda", "United Republic of Tanzania", "South Sudan", "Burundi", "Democratic Republic of the Congo", "Somalia")) %>%
  group_by(`Geographic area`) %>%
  slice_max(order_by = Year, n = 1) %>%
  ungroup()

#  Filter to Sex == "Total" and Wealth Quintile == "Total"
neonatal_latest <- neonatal_latest %>%
  filter(Sex == "Total", `Wealth Quintile` == "Total")

neonatal_latest <- neonatal_latest %>%
  group_by(`Geographic area`, Year) %>%
  summarise(Estimate = median(`Observation Value`, na.rm = TRUE)) %>%
  ungroup()

#rename Tanzania
neonatal_latest <- neonatal_latest %>%
  mutate(`Geographic area` = ifelse(`Geographic area` == "United Republic of Tanzania", "Tanzania", `Geographic area`))

mortality_latest <- mortality_latest %>%
  mutate(`Geographic area` = ifelse(`Geographic area` == "United Republic of Tanzania", "Tanzania", `Geographic area`))

# Read each shapefile
kenya <- st_read("shapefiles/gadm41_KEN_0.shp")
uganda <- st_read("shapefiles/gadm41_UGA_0.shp")
#tanzania <- st_read("shapefiles/gadm41_TZA_0.shp")
#rwanda <- st_read("shapefiles/gadm41_RWA_0.shp")
#burundi <- st_read("shapefiles/gadm41_BDI_0.shp")
#south_sudan <- st_read("shapefiles/gadm41_SSD_0.shp")
#drc <- st_read("shapefiles/gadm41_COD_0.shp")
#somalia <- st_read("shapefiles/gadm41_SOM_0.shp")

# Merge into one object
eac_countries1 <- bind_rows(
  kenya, uganda
)
#, tanzania, rwanda, burundi, south_sudan, drc, somalia
#Merge the data
eac_map_neonatal <- left_join(eac_countries1, neonatal_latest, by = c("COUNTRY" = "Geographic area"))
eac_map_mortality <- left_join(eac_countries1, mortality_latest, by = c("COUNTRY" = "Geographic area"))

#Visual map for Under 5 mortality rate
ggplot(data = eac_map_mortality) +
  geom_sf(aes(fill = Estimate), color = "black") +
  scale_fill_viridis_c(option = "magma", name = "Under-5 Mortality Rate") +
  theme_minimal() +
  labs(
    title = "Latest Under-5 Mortality Rate in East African Community",
  )
```

**Visual Map for latest estimate of Neonatal mortality rates**

```{r,echo=TRUE, message=FALSE, warning=FALSE,fig.width=10, fig.height=6}
#Visual map for neonatal mortality rate
ggplot(data = eac_map_neonatal) +
  geom_sf(aes(fill = Estimate), color = "black") +
  scale_fill_viridis_c(option = "plasma", name = "Neonatal Mortality Rate") +
  theme_minimal() +
  labs(
    title = "Latest Neonatal Mortality Rate in East African Community",
  
  )

```

**Average trends in the mortality rates over time**

```{r,echo=TRUE, message=FALSE, warning=FALSE,fig.width=10, fig.height=6}
#Average trend lines plot
#  calculate the average under-5 mortality per year
mortality_avg <- under_five_eac %>%
  group_by(Year) %>%
  summarize(Average_Estimate = mean(`Observation Value`, na.rm = TRUE))

# Plot
ggplot() +
  geom_line(data = mortality_avg, aes(x = Year, y = Average_Estimate), color = "black", size = 0.8) +
  geom_point(data = under_five_eac, aes(x = Year, y = `Observation Value`, color = `Geographic area`), size = 1) +
  theme_minimal() +
  labs(
    title = "Under-5 Mortality Rate Trends (EAC Countries)",
    x = "Year",
    y = "Mortality Rate (per 1000 live births)",
    color = "Country",
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),  
    axis.text = element_text(size = 12),   
    legend.title = element_text(size = 12), 
    legend.text = element_text(size = 12)  
  )

```

```{r,echo=TRUE, message=FALSE, warning=FALSE,fig.width=10, fig.height=6}
# Calculate average neonatal mortality per year
neonatal_avg <- neonatal_eac %>%
  group_by(Year) %>%
  summarize(Average_Estimate = mean(`Observation Value`, na.rm = TRUE))

# Plot
ggplot() +
  geom_line(data = neonatal_avg, aes(x = Year, y = Average_Estimate), color = "black", size = 1.5) +
  geom_point(data = neonatal_eac, aes(x = Year, y = `Observation Value`, color = `Geographic area`), size = 2) +
  theme_minimal() +
  labs(
    title = "Neonatal Mortality Rate Trends (EAC Countries)",
    x = "Year",
    y = "Mortality Rate (per 1000 live births)",
    color = "Country"
  ) +
   theme(
    plot.title = element_text(hjust = 0.5),  
    axis.text = element_text(size = 12),   
    legend.title = element_text(size = 12), 
    legend.text = element_text(size = 12)  
  )
```

**Country with highest under 5 Mortality rates**

```{r}
# For Under-5 mortality
mortality_latest %>%
  arrange(desc(Estimate)) %>%
  select(`Geographic area`, Estimate) %>%
  head(1)
```

**Country with highest Neonatal mortality**

```{r}
# For Neonatal mortality
neonatal_latest %>%
  arrange(desc(Estimate)) %>%
  select(`Geographic area`, Estimate) %>%
  head(1)
```
